---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

set.seed(4394)
```

# rankGap

<!-- badges: start -->
<!-- badges: end -->

This package contains functions for computing rank-gap and related statistics,
and plotting the results.

These statistics provide a convenient way to represent greater-than-expected
concordance among items ranked highly across a number of lists.
They were developed to understand "overlap" and in genomics and high throughput
biology, but may be useful more broadly.

## Installation

You can install the development version of `rankGap` from
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("l-d-s/rankGap")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(rankGap)

rank_gap(
  rnorm(1e3), rnorm(1e3)
) |> hist()
```

### Citation

The method was developed in Leon Di Stefano's PhD thesis:

```
@thesis{stefanoTamingInteractionsGenomics2024,
  title = {Taming {{Interactions}} in {{Genomics}} and {{Clinical Trials}}},
  author = {Di Stefano, Leon S},
  date = {2024-09-04},
  institution = {Johns Hopkins University},
  url = {https://jscholarship.library.jhu.edu/handle/1774.2/70173},
  urldate = {2025-01-08},
  abstract = {This dissertation develops statistical methods to address two challenges in contemporary biomedical research. The first is understanding shared mechanisms underlying related diseases using high- throughput molecular data, motivated by a study of gene expression changes in a group of rare genetic disorders called Mendelian Disorders of the Epigenetic Machinery or MDEMs (Luperchio et al., 2021). We develop a simple approach to assessing the degree of overlap among the disorders based on rescaled differences in ranks of each gene across conditions, which we call “rank-gap” statistics. Rank-gap statistics have p-value-like properties that highlight concordance among genes ranked highly in at least one condition, and we argue that the number of smaller than expected rank-gap statistics estimates a meaningful overlap metric in the limit of small measurement error. Rank-gap statistics also possess three- and higher-way analogues. We compare our proposal with other rank-based overlap methods in the literature, and evaluate a diverse group of related methods in a simulation study. Our approach provides a simple complement to the predominant practice in genomics of forming Venn diagrams of statistically significant genes. The second challenge is understanding how treatment effects vary among different types of patients using data from clinical trials. The standard approach—one-at-a-time subgroup or interaction analysis of candidate covariates—suffers from difficulties of inference and interpretation. We develop the proportional interaction model, a parsimonious extension of additive regression adjustment also considered by Follmann and Proschan (1999) and Kovalchik, Varadhan, and Weiss (2013) in which benefits and harms from treatment vary as a function of a “score” that also predicts outcomes in each treatment group. An asymptotic analysis enables us to assess the sample sizes required to detect proportional interactions in practice, and we propose a symmetrical re-parametrization of the model that allows for stable single-step inference. We re-analyze a clinical trial of treatments for schizophrenia using our approach, showing how proportional interactions can be incorporated into a traditional regression model-building framework.},
  langid = {american}
}
```

